---
title: "Advanced Plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 5.5, fig.height = 4.2,
                      dev.args = list(bg = "white"))
library(plotit)
```

Techniques beyond the first pipeline: multi-panel figures, relational
layouts, scale control, data-prep recipes, and the ggplot2 escape hatch.

## Multi-panel composition

`compose_*()` assembles **independent** plots (different data, geometries,
scales). `split_*()` facets **one** dataset. All composers return a
`plotit_composite` that continues into `label_*()`, `style()`, and `export()`.

| Function | Layout |
|:---|:---|
| `compose_grid()` | rows × columns |
| `compose_inset()` | floating overlay |
| `compose_marginal()` | scatter + marginal distributions |
| `compose_annot()` | base plot + annotation strips (e.g. dendrogram) |

```{r}
p1 <- iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point(alpha = 0.6)
p2 <- iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_boxplot()
compose_grid(p1, p2, ncol = 2, tag_levels = "A") |>
  label_title("Iris dashboard")
```

```{r}
main <- iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point()
top <- iris |>
  plotit(encode(x = Sepal.Width, fill = Species)) |>
  mark_histogram(bins = 15, alpha = 0.5)
right <- iris |>
  plotit(encode(x = Sepal.Length, fill = Species)) |>
  mark_histogram(bins = 15, alpha = 0.5) |>
  project_cartesian(flip = TRUE)
compose_marginal(main, top, right)
```

```{r}
compose_inset(p1, p2, left = 0.6, bottom = 0.6, right = 0.95, top = 0.95)
```

## Relational layouts

A layout is a **data transform**, not a layer. `as_graph()` normalises
relations into a `plotit_graph`; `layout_*()` bakes coordinates into its
tables; marks render sub-tables with `data = ~table`.

```
edges / matrix / tree ──as_graph()──▶ plotit_graph ──layout_*()──▶ x/y columns
                                              │
                    plotit(graph) ──▶ mark_*(data = ~nodes | ~edges | ~ribbons)
```

```{r}
flows <- data.frame(
  source = c("A", "A", "B", "B", "C"),
  target = c("B", "C", "C", "D", "D"),
  value = c(10, 5, 8, 3, 6)
)
# Sugar path
flows |>
  plotit(encode(source = source, target = target, value = value, fill = source)) |>
  mark_sankey()
```

```{r}
# Explicit layout path
as_graph(flows) |>
  plotit() |>
  layout_sankey() |>
  mark_polygon(data = ~ribbons) |>
  mark_rect(data = ~nodes)
```

```{r}
edges <- data.frame(source = c("a", "a", "b", "c"), target = c("b", "c", "d", "d"))
nodes <- data.frame(id = c("a", "b", "c", "d"), type = c("x", "y", "x", "y"))
as_graph(edges, nodes = nodes) |>
  plotit() |>
  layout_force(seed = 4) |>
  mark_rule(data = ~edges) |>
  mark_point(data = ~nodes)
```

| Layout | Engine | Notes |
|:---|:---|:---|
| `layout_force` | Fruchterman–Reingold | `seed` required |
| `layout_circle` | trigonometric | `order_by` |
| `layout_tree` | leaf-order walk | `direction` |
| `layout_dendrogram` | hclust heights | `direction` |
| `layout_chord` | sector + Bézier ribbons | deterministic |
| `layout_sankey` | layered + Bézier ribbons | deterministic |
| `layout_treemap` | squarify | hierarchical table |

Relational sugar marks (`mark_sankey`, `mark_treemap`, `mark_network`,
`mark_chord`) call the same engines. Prefer sugar for one-liners; drop to
`layout_*` when you need custom marks on sub-tables.

## Scales in depth

All `scale_*()` functions share `name`, `trans`, `limits`, `range`,
`breaks`, `labels`.

| `trans` | Effect | Typical aesthetic |
|:---|:---|:---|
| `"identity"` | linear (default for x/y) | position |
| `"log"`, `"log10"`, `"log2"`, `"sqrt"` | transform data | x, y |
| `"reverse"` | flip order | most |
| `"discrete"` | treat as categories | colour, shape, … |
| `"binned"` | bin then discretise | colour, fill, size |

`range` is the **visual output domain** (Vega-aligned):

- colour/fill: scheme name (`"viridis"`, `"brewer"`, `"friendly"`, `"rdbu"`, …) or colour vector
- size / alpha: numeric bounds (`c(1, 6)`, `c(0.1, 1)`)
- shape / linetype: shape codes / linetype names

```{r}
mtcars |>
  plotit(encode(x = wt, y = mpg, colour = hp)) |>
  mark_point() |>
  scale_x(trans = "log10") |>
  scale_color(range = "rdbu", mid = 0)
```

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Sepal.Length)) |>
  mark_point() |>
  scale_color(trans = "binned", n_bins = 5)
```

Date/POSIXct columns auto-route to a date axis — do **not** force
`trans = "identity"` on them.

## Labels, themes, export

Three-parameter protocol: `reset` > `hide` > `text`.

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point() |>
  label_title("Iris Measurements") |>
  label_subtitle("Anderson's Iris Data") |>
  label_caption("Source: R.A. Fisher, 1936") |>
  label_axis("Sepal Width (cm)", aes = "x") |>
  label_legend("Species", aes = "colour") |>
  style(base_size = 12)
```

```r
export(p, "figure.pdf", width = 8, height = 5, dpi = 300)
```

## The ggplot2 escape hatch

`add_ggplot()` appends any ggplot2 layer, guide, or theme while returning a
plotit object so the pipe continues.

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length)) |>
  mark_point() |>
  add_ggplot(ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "#E15759"))
```

Prefer the verb API; use `add_ggplot()` for features plotit has not wrapped.

## Extending plotit

```r
make_mark("mark_spoke", ggplot2::geom_spoke)

style_dark <- make_theme(
  "style_dark",
  plot.background = ggplot2::element_rect(fill = "#1a1a1a"),
  text = ggplot2::element_text(colour = "white")
)
```

Custom marks share the same registration path as built-ins (`position`,
`rasterize`, defaults).

## Data-prep recipes

plotit consumes tidy tables; reshape with dplyr/tidyr before the pipe.

### Rank / bump chart

```{r}
set.seed(7)
df <- data.frame(
  year = rep(c(2018, 2022), each = 4),
  team = rep(letters[1:4], 2),
  score = c(70, 80, 90, 60, 75, 85, 95, 65)
)
df |>
  dplyr::group_by(year) |>
  dplyr::mutate(rank = dplyr::row_number(dplyr::desc(score))) |>
  dplyr::ungroup() |>
  plotit(encode(x = year, y = rank, group = team, colour = team)) |>
  mark_line() |>
  mark_point()
```

### Waterfall

```{r}
steps <- data.frame(
  item = paste0("s", 1:5),
  delta = c(100, -40, 30, -20, 50)
)
steps$end <- cumsum(steps$delta)
steps$start <- steps$end - steps$delta
steps$dir <- ifelse(steps$delta >= 0, "up", "down")
steps |>
  plotit(encode(x = item, y = end, fill = dir)) |>
  mark_rect(mapping = encode(xmin = item, xmax = item, ymin = start, ymax = end)) |>
  mark_rule(yintercept = 0, colour = "#4E79A7")
```

### Mean ± SE summary

```{r}
se <- function(x) stats::sd(x) / sqrt(length(x))
grpm <- data.frame(
  Species = levels(iris$Species),
  mean = tapply(iris$Sepal.Length, iris$Species, mean),
  sem = tapply(iris$Sepal.Length, iris$Species, se)
)
grpm |>
  plotit(encode(x = Species, y = mean, ymin = mean - sem, ymax = mean + sem)) |>
  mark_point(size = 3) |>
  mark_errorbar(width = 0.2)
```

### Group envelope

```{r}
set.seed(7)
pts <- data.frame(
  x = rnorm(60), y = rnorm(60),
  g = rep(letters[1:3], each = 20)
)
pts |>
  plotit(encode(x = x, y = y, colour = g)) |>
  mark_point(alpha = 0.6) |>
  mark_encircle(shape = "ellipse")
```

## Next

- [Gallery](visualizing-data.html) — chart families by intent
- [API](api.html) — the grammar as a system
- [Design Goals](design-goals.html) — why the grammar looks this way
