---
title: "Design Goals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 5.2, fig.height = 4.0,
                      dev.args = list(bg = "white"))
library(plotit)
```

## Composition first

If a visual can be expressed by combining `mark_*` + `project_*` + `scale_*` +
`split_*` in one pipeline, plotit does **not** add a mark for it. High-value
combinations ship as recipes in the [Gallery](visualizing-data.html). A new
mark is only added when the visual cannot be expressed in a reasonable
pipeline (external layout algorithms, novel data encodings).

```{r}
# Pie is a recipe: mark_bar + project_polar — not mark_arc
data.frame(cat = c("A", "B", "C"), n = c(40, 35, 25)) |>
  plotit(encode(x = 1, y = n, fill = cat)) |>
  mark_bar(position = "stack", width = 1) |>
  project_polar(theta = "y", inner_radius = 0.4)
```

## Verb-prefix grammar

Every exported function starts with a verb that names its role: `mark_*`
adds geometry, `scale_*` maps data to visuals, `project_*` changes
coordinates, `split_*` facets, `label_*` sets text, `compose_*` assembles
panels, `layout_*` bakes relational coordinates, `style` themes, `export`
writes files. One verb, one meaning. All of them return a plotit object so
the native pipe `|>` is the only composition operator users need.

## Zero-dependency relational layouts

Sankey, chord, treemap, network, tree and dendrogram layouts are
self-built, deterministic R engines. No external graph library is required —
publication-reproducible by construction (`ggbeeswarm` is the single,
documented exemption).

## Defaults that publish

Canvas size, colour, dodge, and theme are chosen so a first plot is already
report-ready. Users override deliberately (`scale_*`, `style`, `label_*`)
instead of rebuilding chrome from scratch.

## Explicitly out of scope

plotit targets static print/publication output. Interaction (tooltips,
brushing, linking), 3D, animation, and automatic mark selection are out of
scope. Statistical p-values are computed by the user and fed into
`mark_significance()` as a table.

## The pipeline is the narrative

```{r}
ggplot2::mpg |>
  plotit(encode(x = displ, y = hwy, colour = class)) |>
  mark_point(size = 2, alpha = 0.7) |>
  scale_color(range = "friendly") |>
  label_title("Fuel economy by class")
```

## Next

- [Get Started](plotit.html) — first pipelines
- [Gallery](visualizing-data.html) — chart families by intent
- [API](api.html) — verb families as a system
