---
title: "Getting Started with plotit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with plotit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**plotit** is a declarative plotting package built on ggplot2. It wraps ggplot2 with a
**verb-prefix API** — every function starts with a verb that tells you what it does:
`mark_*()` adds marks, `scale_*()` controls scales, `label_*()` sets labels.

All functions return a plotit object, so you can chain them with `|>`:

```{r}
library(plotit)

iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point(size = 2, alpha = 0.7) |>
  scale_color(range = "viridis") |>
  label_title("Iris Sepal Dimensions") |>
  label_axis(text = "Sepal Width", aes = "x") |>
  label_axis(text = "Sepal Length", aes = "y")
```

## Pipeline Grammar

Every plotit pipeline follows the same six-step grammar:

```
data |> plotit(encode(...)) |> mark_*() |> scale_*() |> label_*() |> style() |> export()
```

| Step   | Function      | Job                                |
|:-------|:--------------|:-----------------------------------|
| 1. Init | `plotit()`   | Create plot with data & aesthetics |
| 2. Mark | `mark_*()`   | Add geometric layers               |
| 3. Scale | `scale_*()`  | Control data-to-visual mapping     |
| 4. Label | `label_*()`  | Set titles, axis labels, legends   |
| 5. Style | `style()`    | Apply a ggplot2 theme              |
| 6. Export | `export()`  | Render to file                     |

## Function Families

### `mark_*()` — Geometric Layers

Six mark functions add visual elements to your plot. All share a unified signature:
`mapping`, `data`, `position`, `rasterize`, and `...` forwarded to the underlying geom.

```{r}
# Scatter plot
mtcars |>
  plotit(encode(x = wt, y = mpg, colour = factor(cyl))) |>
  mark_point(size = 3, alpha = 0.8)
```

```{r}
# Bar chart <U+2014> auto-detects geom_col vs geom_bar
iris |>
  plotit(encode(x = Species, y = Sepal.Length)) |>
  mark_bar()
```

```{r}
# Boxplot
iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_boxplot()
```

```{r}
# Histogram
iris |>
  plotit(encode(x = Sepal.Length, fill = Species)) |>
  mark_histogram(bins = 20, alpha = 0.5)
```

### `scale_*()` — Data-to-Visual Mapping

Eight scale functions, all with identical parameters: `name`, `trans`, `limits`,
`range`, `breaks`, `labels`, and `...`.

```{r}
mtcars |>
  plotit(encode(x = wt, y = mpg, colour = hp, size = hp)) |>
  mark_point(alpha = 0.7) |>
  scale_color(range = "viridis") |>
  scale_x(trans = "log10") |>
  scale_size(range = c(0.5, 8))
```

The `range` parameter accepts colour scheme names (`"viridis"`, `"brewer"`, `"hue"`)
or custom vectors:

```{r}
mtcars |>
  plotit(encode(x = wt, y = mpg, colour = factor(cyl))) |>
  mark_point(size = 3) |>
  scale_color(range = c("#E41A1C", "#377EB8", "#4DAF4A"))
```

### `label_*()` — Text Labels

Five label functions use a three-parameter protocol:

| Call                            | Behaviour                              |
|:--------------------------------|:---------------------------------------|
| `label_*(text = "str")`         | Set custom text                        |
| `label_*(hide = TRUE)`          | Remove element and its space           |
| `label_*(reset = TRUE)`         | Restore variable name or remove text   |

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point() |>
  scale_color(range = "brewer") |>
  label_title("Iris Measurements") |>
  label_subtitle("Anderson's Iris Data") |>
  label_caption("Source: R.A. Fisher, 1936") |>
  label_axis("Sepal Width (cm)", aes = "x") |>
  label_axis("Sepal Length (cm)", aes = "y") |>
  label_legend("Species", aes = "colour")
```

### `project_*()` — Coordinate Systems

```{r}
# Flipped coordinates
iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_boxplot() |>
  project_cartesian(flip = TRUE)
```

```{r}
# Zoom via xlim/ylim
mtcars |>
  plotit(encode(x = wt, y = mpg)) |>
  mark_point() |>
  project_cartesian(xlim = c(2, 4), ylim = c(15, 25))
```

### `split_*()` — Facets

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length)) |>
  mark_point() |>
  split_wrap(Species, ncol = 3)
```

### `style()` — Themes

```{r}
iris |>
  plotit(encode(x = Sepal.Width, y = Sepal.Length, colour = Species)) |>
  mark_point() |>
  scale_color(range = "viridis") |>
  style(ggplot2::theme_minimal(base_size = 14))
```

## Export

```r
p <- mtcars |>
  plotit(encode(x = wt, y = mpg, colour = factor(cyl))) |>
  mark_point(size = 2) |>
  label_title("Fuel Economy")

export(p, "mtcars_plot.png", width = 8, height = 5, dpi = 300)
```
