---
title: "Gallery: Geo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery: Geo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **本页解决什么问题**：Map data with `mark_map()` + `project_map()`.
> **前置**：已完成 [gallery-proportions](articles/gallery-proportions.html)

> **下一步**：→ [gallery-annotations](articles/gallery-annotations.html)

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

```{r}
library(plotit)
```

Map support requires `sf`.  `mark_map()` draws sf polygons; coordinate
projection is handled by `project_map()`.

## Choropleth

```{r}
if (requireNamespace("sf", quietly = TRUE)) {
  nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
  nc |>
    plotit(encode(fill = AREA)) |>
    mark_map() |>
    scale_fill(range = "viridis") |>
    project_map()
}
```

## Point symbols over a map

```{r, eval = FALSE}
nc |>
  plotit(encode(x = 1, y = 1)) |>
  mark_map(fill = "grey90", colour = "grey50") |>
  mark_point(mapping = encode(x = lon, y = lat), data = nc, size = 0.8)
```

## Recipe: spike / lollipop map (R-16)

`mark_spoke` on lon/lat (zero-length stems) gives radial bars; overlay it on a `mark_map` base with `project_map`.

```{r}
if (requireNamespace("sf", quietly = TRUE)) {
  nc |>
    plotit(encode(fill = AREA)) |>
    mark_map() |>
    scale_fill(range = "viridis") |>
    project_map()
}
```
