---
title: "Gallery: Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery: Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **本页解决什么问题**：Explore one or two variables: histograms, density, ECDF, QQ, 2D bins.
> **前置**：已完成 [gallery-comparisons](articles/gallery-comparisons.html)

> **下一步**：→ [gallery-relationships](articles/gallery-relationships.html)
```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5, out.width = "72%")
library(plotit)
```

Distributions describe how values spread out. plotit's distribution marks reveal
shape, tails, and relationships — from histograms and densities to ECDFs, QQ
plots, and bivariate binning.

## Histograms

### Histogram with 30 bins

`mark_histogram()` bins a continuous variable and counts observations per bin.
The `bins` argument controls the resolution.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_histogram(bins = 30)
```

### Histograms by group

Filling by a categorical variable and lowering alpha lets several histograms
overlap on one panel. A small subset of `mpg` classes keeps the comparison
readable.

```{r}
mpg2 <- ggplot2::mpg[ggplot2::mpg$class %in% c("compact", "suv", "pickup", "subcompact"), ]
mpg2 |>
  plotit(encode(x = hwy, fill = class)) |>
  mark_histogram(bins = 20, alpha = 0.5)
```

## Densities

### Density with custom bandwidth

`mark_density()` estimates a smooth kernel density. The bandwidth `bw` (passed
through `...`) trades smoothness against detail — smaller values follow the
data more closely.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_density(bw = 0.15, alpha = 0.6)
```

### Density by group

Coloured density curves overlay several groups for direct shape comparison.

```{r}
iris |>
  plotit(encode(x = Sepal.Length, colour = Species)) |>
  mark_density()
```

### Histogram and density overlay

Rescale the histogram with `after_stat(density)` so the bars and the density
curve share the same y scale.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_histogram(mapping = encode(y = ggplot2::after_stat(density)), bins = 30, alpha = 0.5) |>
  mark_density(bw = 0.15)
```

## Cumulative Distributions

### Empirical CDF

`mark_ecdf()` draws the empirical cumulative distribution function as a step
with no binning to choose — every point is represented exactly.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_ecdf()
```

## Quantile-Quantile Plots

### QQ plot with normal reference

`mark_qq()` plots sample quantiles against the theoretical quantiles of a
distribution; `mark_qq_line()` adds the fitted reference line. Deviations from
the line signal departures from normality.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_qq() |>
  mark_qq_line()
```

### QQ plot with exponential reference

Pass `distribution` to check against any theoretical distribution — here
`"exp"` for data simulated from an exponential. Both the points and the line
must use the same distribution.

```{r}
set.seed(42)
rex <- data.frame(value = rexp(200))
rex |>
  plotit(encode(x = value)) |>
  mark_qq(distribution = "exp") |>
  mark_qq_line(distribution = "exp")
```

## Bivariate Distributions

### 2D density contours

`mark_density_2d()` draws contour lines of a two-dimensional kernel density —
here the classic `faithful` eruptions/waiting relationship.

```{r}
faithful |>
  plotit(encode(x = eruptions, y = waiting)) |>
  mark_density_2d()
```

### Filled 2D density

With `filled = TRUE` the contour bands are filled, giving a topographic view of
where the density is highest.

```{r}
faithful |>
  plotit(encode(x = eruptions, y = waiting)) |>
  mark_density_2d(filled = TRUE, bins = 10)
```

### Bin 2D

`mark_bin2d()` tiles the plane into a rectangular grid of counts — a clean way
to visualise overplotting in dense data.

```{r}
set.seed(1)
dmid <- ggplot2::diamonds[ggplot2::diamonds$carat < 2, ]
dmid |>
  plotit(encode(x = carat, y = price)) |>
  mark_bin2d(bins = 20)
```

### Hexbin

`mark_hex()` bins into hexagons instead of squares, which avoids the visual bias
of axis-aligned rectangles.

```{r}
dmid |>
  plotit(encode(x = carat, y = price)) |>
  mark_hex(bins = 30)
```

## Distribution Details

### Density with rug

`mark_rug()` adds a tick per observation along the axis, recovering the exact
data positions that a smooth density can hide.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_density() |>
  mark_rug(sides = "b", color = "grey30")
```

### Faceted histograms

`split_wrap()` repeats the histogram for each group in its own panel, avoiding
the overlap of stacked or dodged bins.

```{r}
ggplot2::mpg |>
  plotit(encode(x = hwy)) |>
  mark_histogram(bins = 20) |>
  split_wrap(drv, ncol = 3)
```

### Mean with standard error

Summarise groups into means with error bars for the standard error of the mean.
`mark_errorbar()` takes `ymin`/`ymax` alongside `y`; the point shows the mean.

```{r}
se <- function(x) sd(x) / sqrt(length(x))
grpm <- data.frame(
  Species = levels(iris$Species),
  mean = tapply(iris$Sepal.Length, iris$Species, mean),
  sem = tapply(iris$Sepal.Length, iris$Species, se)
)
grpm |>
  plotit(encode(x = Species, y = mean, ymin = mean - sem, ymax = mean + sem)) |>
  mark_point(size = 3) |>
  mark_errorbar(width = 0.2)
```
