---
title: "Gallery: Comparing Groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery: Comparing Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5, out.width = "72%")
library(plotit)
```

Group comparisons are the bread and butter of statistical graphics: bars, boxes,
violins, and cumulative curves all place summaries or distributions side by side.

## Bars

### Counts by class

Count bars tally the rows per category — map only `x` and `mark_bar()` knows to
count. Here it shows how many cars of each `class` are in the `ggplot2::mpg`
data.

```{r}
ggplot2::mpg |>
  plotit(encode(x = class)) |>
  mark_bar()
```

### Value bars

Map `y` as well and the bars read their heights directly from a data column.
Inline data frames are a convenient source for small, hand-built figures.

```{r}
dfv <- data.frame(cat = c("A", "B", "C", "D"), val = c(12, 7, 19, 5))
dfv |>
  plotit(encode(x = cat, y = val)) |>
  mark_bar()
```

### Stacked bars

Mapping a second categorical variable to `fill` and stacking shows the
composition of every group at a glance.

```{r}
ggplot2::mpg |>
  plotit(encode(x = class, fill = drv)) |>
  mark_bar(position = "stack")
```

### Filled bars

With `position = "fill"` every stack is scaled to a constant height, turning the
comparison into group shares that sum to one.

```{r}
ggplot2::mpg |>
  plotit(encode(x = class, fill = drv)) |>
  mark_bar(position = "fill")
```

### Flipped bars

Long category labels read better along the horizontal axis — flip the
coordinates with `project_cartesian(flip = TRUE)`.

```{r}
ggplot2::mpg |>
  plotit(encode(x = class)) |>
  mark_bar() |>
  project_cartesian(flip = TRUE)
```

### Grouped bars

Colour by a second group and dodge the bars side by side. Value bars need
pre-aggregated data, so summarise with `aggregate()` first.

```{r}
mpg_grp <- aggregate(hwy ~ class + drv, data = ggplot2::mpg, FUN = mean)
mpg_grp |>
  plotit(encode(x = class, y = hwy, fill = drv)) |>
  mark_bar(position = "dodge")
```

## Lollipop and Dumbbell

### Lollipop chart

`mark_lollipop()` anchors each point with a stem at zero, giving bar-like
rankings without the visual weight of a filled rectangle.

```{r}
dfl <- data.frame(cat = LETTERS[1:6], val = c(3, 7, 2, 9, 5, 6))
dfl |>
  plotit(encode(x = cat, y = val)) |>
  mark_lollipop()
```

### Dumbbell chart

`mark_dumbbell()` connects paired before/after values with a line. Encode the
start in `y` and the end in `yend`; each row is one comparison.

```{r}
dfd <- data.frame(
  item = c("A", "B", "C", "D", "E"),
  before = c(3, 5, 2, 8, 4),
  after = c(7, 6, 5, 10, 6)
)
dfd |>
  plotit(encode(x = item, y = before, yend = after)) |>
  mark_dumbbell()
```

## Boxplots

### Boxplot by group

`mark_boxplot()` summarises a distribution with quartiles and outliers. Fill by
the grouping variable so each box is visually distinct.

```{r}
iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_boxplot()
```

### Grouped boxplots

Two grouping variables on the same axes make pairwise comparisons easy — engine
`cyl` on `x`, transmission `am` mapped to `fill`.

```{r}
mtcars |>
  plotit(encode(x = factor(cyl), y = mpg, fill = factor(am))) |>
  mark_boxplot()
```

## Violins and Strips

### Violin plot

A violin shows the full density shape instead of just the quartiles.
`draw_quantiles = 0.5` marks the median with a line across each body.

```{r}
iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_violin(draw_quantiles = 0.5)
```

### Beeswarm

`mark_beeswarm()` packs every point without overlap, preserving each observation
while showing where the data are densest.

```{r}
iris |>
  plotit(encode(x = Species, y = Sepal.Length, colour = Species)) |>
  mark_beeswarm()
```

### Strip plot

A strip plot jitters points horizontally along each group. `set.seed()` keeps
the jitter reproducible across renders.

```{r}
set.seed(42)
iris |>
  plotit(encode(x = Species, y = Sepal.Length, colour = Species)) |>
  mark_point(position = "jitter", alpha = 0.5, size = 1.5)
```

### Boxplot with jitter overlay

Layer a boxplot and jittered points together for the classic raw-data-plus-
summary view. Suppress the boxplot's own outliers so points are not doubled.

```{r}
set.seed(42)
iris |>
  plotit(encode(x = Species, y = Sepal.Length, fill = Species)) |>
  mark_boxplot(outlier.shape = NA) |>
  mark_point(mapping = encode(colour = Species), position = "jitter", alpha = 0.4, size = 1)
```

## Cumulative Distributions

### Single ECDF

`mark_ecdf()` draws the empirical cumulative distribution as a step — every
observation is represented exactly, with no binning parameter to choose.

```{r}
faithful |>
  plotit(encode(x = eruptions)) |>
  mark_ecdf()
```

### ECDF by group

Colouring the ECDF by a group compares entire distributions at once: shifts,
spreads, and tail behaviour are all visible.

```{r}
ec <- data.frame(
  value = c(iris$Sepal.Length, iris$Petal.Length),
  part = rep(c("Sepal", "Petal"), each = 150)
)
ec |>
  plotit(encode(x = value, colour = part)) |>
  mark_ecdf()
```

## Distributions Across Groups

### Side-by-side histograms

Dodging histograms by a fill group aligns bins between categories, making shape
comparisons direct.

```{r}
iris |>
  plotit(encode(x = Sepal.Length, fill = Species)) |>
  mark_histogram(position = "dodge", bins = 20)
```

### Density by group

Overlapping density curves are the smoothest way to compare several groups on
one panel. Mapping both `fill` and `colour` gives a translucent curve outline.

```{r}
iris |>
  plotit(encode(x = Sepal.Length, fill = Species, colour = Species)) |>
  mark_density(alpha = 0.4)
```
