---
title: "Gallery: Annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery: Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **本页解决什么问题**：Annotate charts: text, labels, significance brackets.
> **前置**：已完成 [gallery-geo](articles/gallery-geo.html)

> **下一步**：→ [gallery-composing](articles/gallery-composing.html)

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

```{r}
library(plotit)
```

## Reference lines and segments

```{r}
ggplot2::mpg |>
  plotit(encode(x = displ, y = hwy)) |>
  mark_point(alpha = 0.5) |>
  mark_rule(yintercept = 25, colour = "#E15759")
```

## Significance brackets (auto-stacked, stage 3)

`mark_significance()` with `comparisons` auto-stacks brackets above the data.

```{r}
set.seed(7)
d <- data.frame(g = rep(c("a", "b", "c"), each = 20), v = rnorm(60, rep(c(0, 0.8, 1.6), each = 20)))
d |>
  plotit(encode(x = g, y = v)) |>
  mark_boxplot() |>
  mark_significance(comparisons = data.frame(
    group1 = c("a", "a", "b"),
    group2 = c("b", "c", "c"),
    label = c("*", "**", "ns")
  ))
```

## Forest plot

```{r}
est <- data.frame(
  study = paste0("s", 1:5),
  est = rnorm(5, 1, 0.3),
  lo = c(0.4, 0.7, 0.9, 0.8, 1.1),
  hi = c(1.6, 1.8, 2.0, 1.9, 2.4)
)
est |>
  plotit(encode(y = study, x = est, xmin = lo, xmax = hi)) |>
  mark_forest(ref = 1)
```

## Group encircle (stage 3: mark_encircle)

```{r}
set.seed(7)
pts <- data.frame(
  x = rnorm(60),
  y = rnorm(60),
  g = rep(letters[1:3], each = 20)
)
pts |>
  plotit(encode(x = x, y = y, colour = g)) |>
  mark_point(alpha = 0.6) |>
  mark_encircle(shape = "hull", expand = 0.02)
```

## Error bars and confidence ribbons (stage 3)

```{r}
set.seed(7)
d <- data.frame(g = rep(letters[1:3], each = 12), v = rnorm(36, rep(c(2, 3, 5), each = 12)))
d |>
  plotit(encode(x = g, y = v)) |>
  mark_errorbar(stat = "mean_ci95", width = 0.3)
```
