---
title: "Design Philosophy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Philosophy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

## Composition first

If a visual can be expressed by combining `mark_*` + `project_*` + `scale_*` +
`split_*` in one pipeline, plotit does **not** add a mark for it.  High-value
combinations are collected as recipes (see the gallery pages); only when a
combination cannot be expressed in a reasonable pipeline (external layout
algorithms, novel data encodings) does a new mark ship.

## Zero-dependency relational charts

Sankey, chord, treemap, network, tree and dendrogram layouts are all
self-built, deterministic R engines.  No external graph library is required —
publication-reproducible by construction (`ggbeeswarm` is the single,
documented exemption).

## A static publishing pipeline

plotit targets print/publication output.  Interaction (tooltips, brushing,
linking), 3D, animation and auto-mark selection are explicitly **out of
scope**; statistical p-values are computed by you and fed into
`mark_significance()` as a table.

```{r}
library(plotit)
```

## The pipeline is the narrative

```{r}
ggplot2::mpg |>
  plotit(encode(x = displ, y = hwy, colour = class)) |>
  mark_point(size = 2, alpha = 0.7) |>
  scale_color(range = "friendly") |>
  label_title("Fuel economy by class")
```